---
title: "data-viz-by-mtr"
author: "Marie Rivers"
date: "8/18/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# General use packages:
library(tidyverse)
library(here)
library(janitor)

# Specifically for plots
library(ggplot2)
library(patchwork)
library(gghighlight)
library(paletteer)
library(ggExtra)
library(ggrepel)
library(ggbeeswarm)

# and for another dataset we'll explore:
library(gapminder)

# Spatial
library(sf) # for working with spatial data in R
```

```{r}
# read in data
lizards <- read_csv(here("data_tidy", "lizards.csv"))

# type `here()` in the console to determine where "here" is
```
Histogram (really bad) of lizard weights
```{r}
# histograms only need one variable
ggplot(data = lizards, aes(x = weight)) +
  geom_histogram(fill = "orange", color = "blue",
                 size = 2,
                 linetype = "dotted")

# Color refers to points and lines. Fill refers to polygons
# Size of polygon refers to the width of the border
```
Scatter Plot
```{r}
# total_length on x-axis; weight on y-axis
# each shape number corresponds a different shape (0-25). some have fill colors. or you can use ascii codes

ggplot(data = lizards, aes(x = total_length, y = weight)) +
  geom_point(shape = 22,
             fill = "yellow",
             color = "red",
             size = 3,
             alpha = 0.7) # transparency: 0 = completelly transparent, 1 = opaque 
```
Bubble plot (take a point and change its size based on a variable)
```{r}
# bubble plot where the color of the points changes based on common_name and the size of points varies based on total_lenth.
# DISCLAIMER: BAD IDEA

ggplot(data = lizards, aes(x = total_length, y = weight)) +
  geom_point(aes(color = common_name, size = total_length),
             shape = 22,
             fill = "black") +
  theme_light()
```
Facet grid vs. facet wrap
#### facet wrap is a bunch of graphs in a row that wraps below if needed for space
```{r}
ggplot(data = lizards, aes(x = total_length, y = weight)) +
  geom_point(aes(color = common_name)) +
  theme_light() +
  facet_wrap(~common_name, ncol = 4) # `~` means "by"
# you can change the number of columns in facet wrap. fills up first row first, then goes onto the next row
```
Facet Grid
```{r}
ggplot(data = lizards, aes(x = total_length, y = weight)) +
  geom_point(aes(color = common_name)) +
  facet_grid(sex ~ tail) # row, then columns
# populates with data at each intersection
```
Getting things in order
(default is alphabetical)
```{r}
# find total lizard counts by common_name
lizard_count <- lizards %>% 
  group_by(common_name) %>% 
  summarise(count = n())

# or
lizard_count2 <- lizards %>% dplyr::count(common_name)
lizard_count_name_tail <- lizards %>%  count(common_name, tail)

# make common_name a factor, and assign each factor a level based on a value in another variable. Add a negative sign infront of the other variable to flip the order.

# `fct_reorder` = factor reorder
ggplot(data = lizard_count, aes(y = fct_reorder(common_name, count), x = count)) +
  geom_col(aes(fill = common_name), show.legend = FALSE) +
  labs(x = "Lizard counts",
       y = "Number of Lizards")
# switched y and x to make it more readable or use `coord_flip()`
```
Try converting common name to an ordered factor outside of ggplot
```{r}
# make common name a factor based on count
common_name_factor <- lizard_count %>% 
  mutate(common_name = fct_reorder(common_name, count))

levels(common_name_factor$common_name)
```
## Axis scale breaks, limits and labels
Scatterplot: total_length (x) vs. weight (y)
```{r}
ggplot(data = lizards, aes(x = total_length, y = weight)) +
  geom_point() +
  scale_x_continuous(breaks = c(0, 10, 50, 100, 500),
                     limits = c(0, 500),
                     expand = c(0, 0)) +
  scale_y_continuous(breaks = seq(from = 0, to = 70, by = 10),
                     limits = c(0, 70),
                     expand = c(0, 0))
# breaks are the tic mark breaks. the vector can be created manually or by sequence
# limits are min and max value
```
Transform the date column to class Date, then find counts of observations by date
```{r}
lizard_obs_count <- lizards %>% # if you attach `lubridate` up top you dont need `lubridate::`
  mutate(date = lubridate::mdy(date)) %>% 
  count(date) %>% 
  mutate(count = n)

class(lizard_obs_count$date)
```

Make a line plot (geom_line()) of date (x) and count (y)
```{r}
# `y` lowercase is 2 diget year, `Y` upper case is a 4 year date
ggplot(data = lizard_obs_count, aes(x = date, y = count)) +
  geom_line() +
  scale_x_date(date_breaks = "3 years",
               date_labels = "%Y")
```
Color Scales
```{r}
ggplot(data = lizards, aes(x = total_length, y = weight)) +
  geom_point(aes(color = weight)) +
  scale_color_gradient(low = "red", high = "navy") 

# scale_color_gradient for 2 colors. scale)color)gradientn for multiple colors
```
```{r}
ggplot(data = lizards, aes(x = total_length, y = weight)) +
  geom_point(aes(color = weight)) +
  scale_color_gradientn(colors = c("orange", "cyan", "chartreuse", "black")) 

```
```{r}
# to set breaks between colors
ggplot(data = lizards, aes(x = total_length, y = weight)) +
  geom_point(aes(color = weight)) +
  scale_color_stepsn(colors = c("green", "blue", "purple"),
                     breaks = c(0, 20, 40)) # if you give it more breaks than colors, it will great a gradient between colors 
```
#### Update a color scheme using a palette in paletteer
Make a horizontal boxplot with common_name on the y-axis, total_length on the x-axis, with color changing based on common_name
```{r}
lizards_fct <- lizards %>% 
  mutate(common_name = fct_reorder(common_name, total_length, .fun = median))

ggplot(data = lizards_fct, aes(y = common_name, x = total_length)) +
  geom_boxplot(aes(fill = common_name), show.legend = FALSE) +
  scale_fill_paletteer_d(palette = "beyonce::X10") #palette = "package::palette"
```
`View(paletts_d_names)` note package name and palette name

```{r}
ggplot(data = lizards, aes(y = common_name, x = total_length)) +
  geom_poiny
  scale_fill_paletteer_d(palette =
```

